\textcolor{red}{Simple} connectome inference using partial correlation from imaging calcium signal

### 2014-06-16 article.tex
Move and reduce introduction paragraph

@@

% Solution
In this paper, we describe a simple and a theoretically grounded approach
based on partial correlation\footnote{Source code is available at
\url{https://github.com/asutera/kaggle-connectomics}.} to infer the neural
network, which is the winning method of the Connectomics challenge
\footnote{\url{http://connectomics.chalearn.org}}.

@@

The raw signal is the calcium fluorescence imaging representing neuron
activity via a capture imaging process. Observing the fluorescent calcium as
an indicator of neuron activity delivers a 2-D view of neuron network at each
capture which only gives a network screenshot at given time points. It implies
obvious issues. Some neurons may be invisible because they are not observed or
they are confounded with another one because of the loss of the third
dimension. The capture sample rate is also a constraint that could make
difficult to detect in which order the neurons fire.  Moreover the activity
indicator is highly reactive to a burst but takes some time to recover its non
active state.  Finally, the difficulty to distinguish direct relationships when
 a network burst happens, defined as a highly synchronous activity
 [Stetter and in Stetter [20,46,47]], is increased by the slow sampling rate.

@@

\begin{figure}%{r}{0.7\linewidth}
\caption{Filtering steps.}\label{wrap-fig:1}
\includegraphics[width=\linewidth]{images/fig_filtering_recut.png}
\end{figure}

@@

% However, the image fluorescence calcium signal is also very noisy. Thus,
% A low-pass filter (Equation \ref{eq:low-pass}) is first applied followed by a
% high-pass filter (Equation \ref{eq:high-pass}), the discrete derivative of
% the signal,
% \begin{align}
% X^\prime_{t,i} &= \sum_{l=-3}^2 c_l^k X_{t+l,i} \forall i, \label{eq:low-pass}\\
% X^{\prime\prime}_{t,i} &= X^{\prime}_{t,i} - X^{\prime}_{t-1,i} \forall i, \label{eq:high-pass}
% \end{align}
% where the $c_l^k$ coefficients were taken in one of the four
% following tuples
% $\left(c_l^1\right)_{l=-3}^2=(0, 0, 1, 1, 1, 1)$,
% $\left(c_l^2\right)_{l=-3}^2=(0, 0, 1, 1, 1, 0)$,
% $\left(c_l^3\right)_{l=-3}^2=(1, 1, 1, 1, 0, 0)$ or
% $\left(c_l^4\right)_{l=-3}^2=(0.4, 0.8, 1, 1, 0, 0)$.

As it can be seen on \textit{fig 1}, the raw signal is very noisy due to light
scattering artifacts that ordinarily affects the quality of the recording [in
Stetter, [6]]. Therefore, a low-pass filter is first applied to get a cleaner
version of the signal where only true variations are kept, i.e. without noise
variations. However, the choice of a low-pass filter is not unique. In the
rest of the paper, we will use two particular filters giving similar results,
the symmetrical median filter, i.e., $x^t_i \leftarrow x^{t-1}_i + x^{t}_i +
x^{t+1}_i$, and an asymmetrical weighted median filter taking into account the
history, i.e., $x^t_i \leftarrow 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i
+ x_{t}^i$.

Considering the short delay for a communication between
neurons, slow events can be neglected to retrieve direct relationships.
Indeed, for a direct relationship, i.e. a neuron firing directly implying another
neuron to fire, the average time is about $1ms$ or $2ms$ while the
fluorescent signal time resolution is $20ms$. In order to find out which neurons
are directly related, it is only necessary to look after the burst signal, or
in other words, whether the signal varies. In our approach, we choose to apply
an asymmetrical discrete derivative : $x^{t,d}_{i} \leftarrow x^{t}_i - x_{t-1}^i$.

@@

Then we apply on the entire signal a non-linear filter based on the current
activity of the systems in order to favor the association measure inversely
proportional to the number of neurons firing together: $ x^{t,w}_i  =
(x^{t,d}_i + 1 )^{1 + \frac{1}{act_t}}$ where $act_t = \sum_{j} x_{t,d}^j$ is
the activity of all neurons at time $t$. When too much neurons are firing
together, i.e. in case of a network burst, it is almost impossible to retrieve
direct relationships: all neurons seem to be directly connected to each other.
By favoring the association measure when the activity is low, we expect to
retrieve a pair of neurons that are really directly connected together. In
concrete terms, as it can be verified on \textit{fig 1}, peak intensity may
change in function of the activity of all neurons but the number and the
locations of peaks is not modified.

@@

\section{Experiments}

% Datasets
To assess our method, we have used four datasets provided by the organizers
of the Connectomics challenge : \textit{normal-1, normal-2, normal-3, normal-4}.
Those datasets are time series of size $T=179500$ of image calcium from simulated
neuronal networks (\cite{stetter2012model}) with $|V|=1000$ neurons. The networks
to inferred have around 15000 edges and are similar to those
used to rank challenge participants. Note that our parameter tuning was mainly
done on the \textit{normal-1} dataset and partly on \textit{normal-4} dataset,
therefore it may justify why these networks are better retrieved despite the
fact that some might be simpler to infer.

@@

For the directory method, only filters defined by Equations
\ref{eq:low-pass}, \ref{eq:high-pass} and \ref{eq:hard-treshold-filter} were
used.

\begin{table}[tbh]
\centering
\caption{TODO}
\begin{tabular}{@{}l *{8}{c}@{}}
\toprule
  & \multicolumn{4}{c}{AUROC} & \multicolumn{4}{c}{AUPRC} \\
\cmidrule(r){2-5}\cmidrule(r){6-9}
Methods $\backslash$ normal- & 1 & 2 & 3 & 4 & 1 & 2 & 3 & 4 \\
\midrule
GENIE3                 & & & & & & & & \\
Directivity            & 0.535 & 0.526 & 0.533 & 0.535 & 0.012 & 0.012 & 0.012 & 0.012 \\
Correlation            & 0.886 & 0.884 & 0.891 & 0.877 & 0.153 & 0.145 & 0.170 & 0.132 \\
GTE                    & 0.890 & 0.893 & 0.894 & 0.873 & 0.171 & 0.174 & 0.197 & 0.142 \\
Partial correlation    & 0.932 & 0.931 & 0.930 & 0.928 & 0.364 & 0.366 & 0.359 & 0.344 \\
Our approach           & 0.943 & 0.942 & 0.942 & 0.939 & 0.403 & 0.404 & 0.398 & 0.388 \\
Our approach + dir     & 0.944 & 0.943 & 0.942 & 0.940 & 0.404 & 0.405 & 0.399 & 0.389 \\
\bottomrule
\end{tabular}
\end{table}

@@

>>>>

The fluorescent calcium image delivers at each acquisition a 2-D view of the
neuron network activities. However, some neurons might not observed or are
confounded with others. The low device sampling rate with respect to neuronal
activity speed might not allow to distinguish neuron activation order,
especially in highly synchronous activity [Stetter and in Stetter [20,46,47]]
of a large part of the network. Moreover the fluorescent concentration is
highly reactive to a neuron activation or a spike, but have a slow decay.

@@

% Solution

In this paper, we describe a simple and a theoretically grounded approach
based on partial correlation\footnote{Source code is available at
\url{https://github.com/asutera/kaggle-connectomics}.} to infer the neural
network, which is the winning method of the Connectomics challenge
\footnote{\url{http://connectomics.chalearn.org}}.

\textcolor{red}{It may be useful to rewritte the following lines but I add it to not forget about a plan (request from AQ)}
In section \ref{sec:filter}, we will describe how we filter the fluorescent calcium imaging signal.
In section \ref{sec:inference}, we will characterize our algorithm from a theoretical point of view. 
In section \ref{sec:results}, we will present some results and give a critical comparision with other inference methods.
In appendix \ref{app:optimized}, we describe a fully tuned version of our algorithm which has won the Connectomics challenge.

@@

@@

\begin{figure}%{r}{0.7\linewidth}
\caption{Filtering steps.}\label{wrap-fig:1}
\includegraphics[width=\linewidth]{images/fig_filtering.pdf}
\end{figure}

@@

As it can be seen on \textit{fig 1}, the raw signal is very noisy due to light
scattering artifacts that ordinarily affects the quality of the recording [in
Stetter, [6]]. Therefore, a low-pass filter is first applied to get a cleaner
version of the signal where only true variations are kept, i.e. without noise
variations. However, the choice of a low-pass filter is not unique. In the
rest of the paper, we will use two particular filters giving similar results,
the symmetrical median filter, i.e., $x^t_i \leftarrow x^{t-1}_i + x^{t}_i +
x^{t+1}_i$, and an asymmetrical weighted median filter taking into account the
history, i.e., $x^t_i \leftarrow 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i
+ x^{t}_i$.

Considering the short delay for a communication between
neurons, slow events can be neglected to retrieve direct relationships.
Indeed, for a direct relationship, i.e. a neuron firing directly implying another
neuron to fire, the average time is about $1ms$ or $2ms$ while the
fluorescent signal time resolution is $20ms$. In order to find out which neurons
are directly related, it is only necessary to look after the burst signal, or
in other words, whether the signal varies. In our approach, we choose to apply
an asymmetrical discrete derivative : $x^{t,d}_{i} \leftarrow x^{t}_i - x^{t-1}_i$.

@@

Then we apply on the entire signal a non-linear filter based on the current
activity of the systems in order to favor the association measure inversely
proportional to the number of neurons firing together: $ x^{t,w}_i  =
(x^{t,d}_i + 1 )^{1 + \frac{1}{act_t}}$ where $act_t = \sum_{j} x^{t,d}_j$ is
the activity of all neurons at time $t$. When too much neurons are firing
together, i.e. in case of a network burst, it is almost impossible to retrieve
direct relationships: all neurons seem to be directly connected to each other.
By favoring the association measure when the activity is low, we expect to
retrieve a pair of neurons that are really directly connected together. In
concrete terms, as it can be verified on \textit{fig 1}, peak intensity may
change in function of the activity of all neurons but the number and the
locations of peaks is not modified.

@@

\section{Experiments}
\label{sec:results}

% Datasets
To assess our method, we have used four datasets provided by the organizers
of the Connectomics challenge : \textit{normal-1, normal-2, normal-3, normal-4}.
Those datasets are time series of size $T=179500$ of image calcium from simulated
neuronal networks (\cite{stetter2012model}) with $|V|=1000$ neurons. The networks
to inferred have around 15000 edges and are similar to those
used to rank challenge participants.

@@

% For the directory method, only filters defined by Equations
% \ref{eq:low-pass}, \ref{eq:high-pass} and \ref{eq:hard-treshold-filter} were
% used.

% Dummy classifier?

% GTE : Explaning the method (default parameters!)

% Explaining the method (thèse de Gilles)
% Genie3 : better in auprc while worse in auroc than GTE. Why?
% Genie3 : "Conditioning" better and better but need a lot of trees
% Correlation : Already give good results

% Partial correlation : Obvious improvements (both metric) because of the conditioning

% Directivity : ?

%Opt: obviously when parameters are tuned/optimized, it's slightly better --> see appendix for a full description of the method
% Avg.: Averaging over the threholds and L-P filters improves a lot both metrics.

GTE is a method which characterizes a relationship between neurons by an
estimation of the generalized transfer entropy (\cite{stetter2012model}). The
goal of this method tries to see if the next time step value is better from
the previous values of this time series or from time series of other neurons.

GENIE3 is a tree-based method which sees the network inference problem as $p$,
the number of variables, independent supervised learning problems (todo cite:
Irrthum?Van Anh?Pierre?). Formally, each sub-problem consists in the
prediction of one variable, the target, from the remaining $p-1$ variables.
This method uses random forests to estimate the target variable and variable
importances can be derived to estimate the adjacency matrix $A$ (see (todo
cite) for a more detailed description of the inference method).

At first sight, it appears on \textit{Table 1} that partial correlation
methods outperform other methods. Two main reasons may explain this method
hierarchy. First, the quality of the infered network seems related to the
conditioning part of the method. In every method except with partial
correlation, indirect links are confounded with direct links.  Secondly, the
gaussian assumption seems verified and thus linear method appears to be
competitive, e.g. the method correlation already gives good results, with a
non-linear method such as GENIE3.

% Discussion about

\begin{table}[tbh]
\centering
\caption{TODO}
\begin{tabular}{@{}l *{8}{c}@{}}
\toprule
  & \multicolumn{4}{c}{AUROC} & \multicolumn{4}{c}{AUPRC} \\
\cmidrule(r){2-5}\cmidrule(r){6-9}
Methods $\backslash$ normal- & 1 & 2 & 3 & 4 & 1 & 2 & 3 & 4 \\
\midrule
GENIE3                 & & & & & & & & \\
Directivity \textcolor{red}{appendix?} & 0.535 & 0.526 & 0.533 & 0.535 & 0.012 & 0.012 & 0.012 & 0.012 \\
Correlation            & 0.886 & 0.884 & 0.891 & 0.877 & 0.153 & 0.145 & 0.170 & 0.132 \\
GTE                    & 0.890 & 0.893 & 0.894 & 0.873 & 0.171 & 0.174 & 0.197 & 0.142 \\
Partial correlation    & 0.930 &  0.927 &  0.926 &  0.923& 0.363  & 0.347 &  0.346 & 0.328 \\
Partial correlation (opt.) & 0.932 & 0.931 & 0.930 & 0.928 & 0.364 & 0.366 & 0.359 & 0.344 \\
Our approach (opt. \& avg.)    & 0.943 & 0.942 & 0.942 & 0.939 & 0.403 & 0.404 & 0.398 & 0.388 \\
Our approach + dir  \textcolor{red}{appendix?}   & 0.944 & 0.943 & 0.942 & 0.940 & 0.404 & 0.405 & 0.399 & 0.389 \\
\bottomrule
\end{tabular}
\end{table}

@@

The algorithm propose in sections \ref{sec:filter} and \ref{sec:inference} has been optimized at its extreme limit to win the challenge. In this section, we will describe the full tuned algorithm and present some results.

Note that our parameter tuning was mainly
done on the \textit{normal-1} dataset and partly on \textit{normal-4} dataset,
therefore it may justify why these networks are better retrieved despite the
fact that some might be simpler to infer.
